---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor

```{r}
library(tidyverse)
#install.packages("janitor")
library(janitor)
```



### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv

```{r}
crimes <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv")
```


### Task 3: Use janitor to clean up the column names

```{r}
crimes <- crimes |> clean_names()
```


### Task 4: Calculate the total crime rate for each year (picking our unit!)

```{r}
crimes |> mutate(total_rate = grand_total/population * 10000)
```



### Task 5: Calculate the violent and property crime rates for each year (same unit)

```{r}
crimes_with_rates <- crimes |> mutate(violent_rate = violent_crime_total/population * 10000, property_rate = property_crime_total/population * 10000)
```



### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations? What might explain some of the results you found?

Looking at the data, the crime rate was highest around the early 1990s, with 1991 and 1992 standing out as some of the highest years. After that, the numbers slowly started going down. The lowest crime rates were in the most recent years, especially 2019 and 2020. To me, this pattern was somewhat expected because I have heard that crime was much higher in the late 1980s and early 1990s. What surprised me was how much the numbers dropped over time. There could be a few reasons for this, such as changes in policing, the economy, technology, and changes in how people report or prevent crime. The data shows that crime did not stay at the same level and changed a lot depending on the time periodthere also could be a lot of crimes that are not reported, like in NIBRS, thats how crimes sre reported. 


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

