---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```




### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```

### Task 3: Calculate the percentage change in each category between Fall 2026 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2026`- `Fall 2021`)/`Fall 2021` * 100)
```



### Task 4: Describe the results

Write a few sentences about what you found in those percentages. What's the story here? What questions do you have?
The percentages have a big change as far as student sustainability and the shuttle bus. The story here is, if student sustaibality is the highest change in percentage, what actually counts as student sustainability; because as of now we dont see much chnages of where out tuition money is going towards. we might see EOS on campus cleaning up outside, fixing the plants and cutting grass but where is the trash cans around the streets. but also looking at the counseling, what is genuinely going on right now. Some of the fees that we are seeing are not necessary because we are not seeing some of the changes going on around campus. Why is the athletics stuck at 399 and why hasnt it gone up at all? 



When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
