---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

The categories include accidents, assaults, stolen vehicles, burglaries, homicides, robberies, sex offenses, and theft. Some categories are much more common than others. For reporting, categories such as homicide, assault, and robbery may be more important because they involve serious or violent crimes, even though they happen less often than accidents or theft. The categories can also make summarizing the data harder because some crimes are divided into smaller groups, such as commercial, residential, and vehicle robberies.

### Task 4: Calculating Averages

Add up the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime using the total number of months, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime %>%
  mutate(
    total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec,
    average_incidents = total_incidents / 12
  )
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The categories with the highest average number of incidents per month are accidents, auto stolen, theft from auto, accident with impound, and theft. The categories with the lowest averages are robbery residential, homicide, B & E other, and sex offense. This shows that more common incidents, such as accidents and theft, happen much more frequently than serious violent crimes. I would have questions about some of the categories with very low numbers because they may be harder to compare with categories that happen much more often.
### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages %>%
  mutate(
    june_change = (Jun - average_incidents) / average_incidents * 100,
    july_change = (Jul - average_incidents) / average_incidents * 100,
    august_change = (Aug - average_incidents) / average_incidents * 100
  )
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

The summer months show different patterns depending on the category of crime. Some categories were higher than their average in June, July, or August, while others were lower than their average. This means that crime did not increase across every category during the summer. The data gives some support to the idea that certain types of crime may be more common during the summer, but the pattern is not consistent enough to say that crime overall is always higher during the summer months.
### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

my lede: Accidents, auto theft, and theft from autos were some of the most common incidents reported in Prince Georges County during 2024.

my nutgrapfh: The 2024 crime data shows that the most common incidents in Prince George's County were not necessarily the most serious crimes. Accidents and different types of theft occurred much more often than crimes such as homicide and robbery. The data also shows that crime patterns changed depending on the category and month, including during the summer. Looking at these differences gives a better understanding of what types of incidents were most common in the county instead of simply looking at the overall number of crimes.
### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
