---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- mutate(md_cities,
  violent_rate_2019 = violent2019 / pop2019 * 1000,
  property_rate_2019 = property2019 / pop2019 * 1000,
  violent_rate_2020 = violent2020 / pop2020 * 1000,
  property_rate_2020 = property2020 / pop2020 * 1000
)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

in 2019, baltimore had the highest violent crime rate at about 18.3 crimes per 1,000 people, while ocean city had the highest property crime rate at about 103.7 per 1,000 people. In 2020, ocean city had the highest violent crime rate at about 24.0 per 1,000 people and also had the highest property crime rate at about 121.2 per 1,000 people. the biggest pattern is that the cities with the highest crime rates were not always the largest cities in Maryland. The story here is that crime rates can look very different when population size is taken into account. Looking only at the total number of crimes could make it seem like larger cities have the most crime, but the rates show that smaller cities can also have much higher crime compared to their populations.
### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- mutate(md_cities_rates,
  violent_rate_change = (violent_rate_2020 - violent_rate_2019) / violent_rate_2019 * 100,
  property_rate_change = (property_rate_2020 - property_rate_2019) / property_rate_2019 * 100
)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

looking at the percentage changes, the outliers are more interesting than the overall pattern. Some cities had very large increases in violent or property crime, while others had decreases of 100%. The story in the data is that crime changed very differently from city to city between 2019 and 2020. these large changes are especially noticeable in smaller cities because even a few more or fewer crimes can create a big percentage change.
### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

my lede: Crime rates changed dramatically across Maryland cities between 2019 and 2020, with some cities seeing major increases while others saw crime drop completely.

my nutgrap: The data shows that there was no single pattern for crime across Maryland. Some cities had large increases in violent or property crime, while other cities had decreases of 100%. The biggest changes often happened in smaller cities, where a few additional or fewer crimes could make a large difference in the percentage rate. Looking at crime rates instead of just the number of crimes helps show how differently crime affected communities across Maryland.
### Task 9: Save Your Work and Put it on GitHub

### Task 10: Percentiles and IQR
```{r}

md_cities_percentiles <- mutate(md_cities_rates,
  violent_percentile = ntile(violent_rate_2020, 100)
)
```
```{r}
summarize(md_cities_percentiles,
  cities_above_90 = sum(violent_percentile > 90)
)
```
```{r}

summarize(md_cities_rates,
  violent_rate_IQR = IQR(violent_rate_2020)
)
```

the results do not change my best story idea from the previous assignment. the percentile results show that only some cities have violent crime rates near the highest levels, while the IQR shows the spread of the middle 50% of cities. this supports my previous idea that crime rates can be very different from city to city and that looking at rates gives a better picture than just looking at the total number of crimes.
Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
